Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-04 _tex/index.tex
Built site for gh-pages

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution
\citep{Andrich_1978}. The mode (mean) of this distribution, representing
the (average) relative separation, is given by the difference between
the modal discriminal processes of the stimuli
\(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
distribution, reflecting the variability in the relative separation, is
calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
discriminal dispersions, while \(\rho\) measures the correlation between
the discriminal processes of the stimuli. This correlation quantifies
the dependence of the judge's perception of the trait in one stimulus on
his perception of the same trait in another.

@@

Notably, the correlation between the discriminal processes, \(\rho\),
plays a pivotal role in determining comparison outcomes by shaping the
distribution of the discriminal difference between the stimuli.
Specifically, as the correlation increases, reflecting a stronger
dependence of the judge's perception of quality in one stimulus on his
perception of the other, the distribution of the discriminal difference
narrows. This narrowing ultimately impacts the area under the curve that
determines the comparison outcome and, consequently, the conclusions
drawn from this outcome.

Figure~\ref{fig-correlation} illustrates how varying correlations
influence the distribution of the discriminal difference for the two
texts depicted in Figure~\ref{fig-discriminal_process}. Since the texts
differ in quality, higher correlations increase the likelihood that the
discriminal difference distinctly favors text B over text A. This is
evident from the larger proportion of area under the curve,
\(P(B > A)\), that lies above zero (shaded gray area in
Figure~\ref{fig-correlation}). Conversely, it is not hard to infer that
if the texts had similar or identical quality levels, higher
correlations would likely reduce the chance that the discriminal
difference distinctly favors one text over the other. This probability
reduction occurs because the distribution of the discriminal difference
would become more narrowly centered around zero (not illustrated).

@@

On the one hand, the assumption of \emph{equal dispersion between
stimuli} asserts that the variability in perceived trait levels remains
consistent across all stimuli. However, Thurstone noted that this
assumption may fail when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these can
introduce greater discrepancies into judgments due to their unique
characteristics. One example of this is the evaluation of creativity. In
this context, for instance, it is plausible that a student completes an
assignment in a highly creative yet unexpected way. Judges, unprepared
for such originality, may experience greater discrepancies in their
judgments of the assignment. As a result, the discriminal difference
associated with this assignment is expected to show a broader
dispersion, reflecting the larger variation in its discriminal process
compared to others. Although direct evidence for this specific scenario
is lacking, {studies such as \citet[chap.~8]{vanDaal_2020} suggest that
misfitting texts tend to exhibit larger dispersion, likely due to
increased confusion among judges regarding the texts.}

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis tests. For
example, these scores have been used to identify `misfit' judges and
stimuli \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018},
detect biases in judges' ratings
\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
or test hypotheses related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

>>>>

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution
\citep{Andrich_1978}. The mode (mean) of this distribution, representing
the (average) relative separation, is given by the difference between
the modal discriminal processes of the stimuli
\(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
distribution, reflecting the variability in the relative separation, is
calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) denote the discriminal
dispersions of the stimuli, while \(\rho\) represents the correlation
between their discriminal processes. This correlation quantifies the
dependence of the judge's perception of the trait in one stimulus on his
perception of the same trait in another.

@@

Notably, the correlation between the discriminal processes, \(\rho\),
plays a pivotal role in determining the comparison outcome by shaping
the distribution of the discriminal difference between the stimuli.
Specifically, as the correlation increases, reflecting a stronger
dependence of the judge's perception of quality in one stimulus on his
perception of the other, the distribution of the discriminal difference
narrows. This narrowing ultimately impacts the area under the curve that
determines the comparison outcome and, consequently, the conclusions
drawn from this outcome.

Figure~\ref{fig-correlation} illustrates how varying correlations
influence the distribution of the discriminal difference for the texts
depicted in Figure~\ref{fig-discriminal_process}. Since the texts differ
in quality, higher correlations increase the likelihood that the
discriminal difference distinctly favors text B over text A. This is
evident from the larger proportion of the area under the curve,
\(P(B > A)\), that lies above zero (shaded gray area in
Figure~\ref{fig-correlation}) compared to curves with lower
correlations. Moreover, although the figure does not illustrate this
scenario, it is reasonable to infer that if the texts had similar or
identical quality levels, higher correlations would likely reduce the
chance of the discriminal difference distinctly favoring one text over
the other. This probability reduction occurs because the distribution of
the discriminal difference would become more narrowly centered around
zero.

@@

On the one hand, the assumption of \emph{equal dispersion between
stimuli} suggests that variability in perceived trait levels is
consistent across all stimuli. However, Thurstone observed that this
assumption may not hold when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
and stimuli can introduce greater judgment discrepancies due to their
unique characteristics. In fact, research on text quality may already
provide evidence of the violation of this assumption, particularly
through the use of misfit statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
Misfit statistics quantify the judgment discrepancies associated with a
given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
In this context, texts identified as ``misfits'' display greater
judgment discrepancies than other texts. This implies that the
discriminal difference associated with these texts shows a broader
dispersion, suggesting that their discriminal processes also exhibit
greater variation than other texts. This reasoning also applies to
``misfit'' judges, whose evaluations reflect substantial deviations due
to the unique or extreme characteristics of the stimuli.

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis tests. For
example, these scores have been used to identify `misfit' judges and
stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
detect biases in judges' ratings
\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
or test hypotheses related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
